---
title: "ggplot2 starters"
author: "Dumky de Wilde"
output: html_document
---

This document is an attempt to set up some basic graph elements for ggplot2 that can be copied and pasted easily when creating new plots and graphs.

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=8, fig.height=6, 
                      fig.align="center", warning=FALSE)
```

Let's start with the basics:

```{r, echo=TRUE}
#libraries:
library("ggplot2")
library("RColorBrewer")
library("stringr")
library("tidyr")
library("reshape2")

p <- ggplot(cars, aes(speed, dist)) +
  geom_point() +
  ggtitle("Cars: speed and distance")

p #And plot it
```

##Background and axes
Now let's make it a little nicer and set up the background and axes

```{r background}

bg_and_axes <-  theme(
  panel.background = element_rect(fill = NA),
  panel.grid.major.y = element_line(colour = "#cccccc", size = 0.3, linetype = "dotted"),
  panel.grid.major.x = element_blank(),
  axis.ticks.y = element_blank(),
  axis.line.x = element_line(color = "black", size = 0.5),
  axis.text = element_text(colour = "#666666"),
  axis.title = element_text(colour = "#999999", face = "italic"),
  axis.text.x = element_text(margin = margin(16,0,12,0)),
  axis.text.y = element_text(margin = margin(0,16,0,12)),
  plot.title = element_text(family = "Avenir", size = rel(2), 
                            margin = margin(0,6,24,6), hjust = 0),
  plot.margin = unit(c(1,1,1,1), "cm"))

p + bg_and_axes
```

##Scales
Set the scales properly:

```{r }
p2 <- p + bg_and_axes + 
  scale_y_continuous(limits = c(0,125), breaks = seq(0,125,25), expand = c(0,0)) +
  scale_x_continuous(limits = c(0,27), breaks = seq(0,25,5), expand = c(0,0))

p2 #And plot our new graph
```

##Bars and colors
Now to some bar charts and colors. I like the spectral color set because it is easy to tell the colors apart, and works for the color blind too, but it still has the feeling of matching colors.

```{r}
p3 <- ggplot(data = diamonds, aes(color, ..count.., fill=cut)) +
  geom_bar() +
  bg_and_axes +
  scale_y_continuous(limits = c(0,12000), expand = c(0,0)) +
  scale_x_discrete(expand = c(0,0)) +
  ggtitle("Diamonds by cut and color")

legend_opts <- theme(
  legend.text = element_text(color = "#666666"),
  legend.title = element_text(color = "#666666", face = "italic"),
  legend.direction = "vertical",
  legend.position = c(0.9,0.9),
  legend.key.size = unit(12, "pt")
  )

p3 +
  legend_opts +
  scale_fill_brewer(palette = "Spectral", direction = -1)
```

Now let's go from stacked bars to non-stacked, and flip 'm over, because I always favor horizontal bars instead of vertical. It makes for easier comparison, especially for longer lists, and improves readability of the label text. Let's take a different data set and you'll see what I mean.

```{r fig.height=7}
#Make sure we have some long label names so you'll see how they wrap.
msleep["full_name"] <- paste(msleep$name, ", also known as: ", 
                             msleep$genus, msleep$order, sep = " ")

#now for the plot.
p4 <- ggplot(msleep[msleep$order == "Primates",], aes(reorder(full_name, sleep_total), sleep_total)) +
  geom_bar(stat="identity", fill=brewer.pal(3, "Set1")[1]) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 25)) +
  scale_y_continuous(expand = c(0,0)) +
  coord_flip()

p4 +
  bg_and_axes +
  ggtitle("Sleep times of primates")
```

